---
title: "01 EDA"
subtitle: "STA120: Introduction to Statistics
         Spring semester 2024
         IMath, University of Zurich, Switzerland"
author: "Reinhard Furrer with contributions of various others"
date: "February 17, 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### R-Code: 
### Loading the lemanHg dataset in R. (Lecture I)
```{r}
Hg.frame <- read.csv('http://user.math.uzh.ch/furrer/download/intro2stat/lemanHg.csv') 
#View(Hg.frame)
str( Hg.frame)      # dataframe with 1 numeric column and 293 observations
head( Hg.frame, 3)  # column name is 'Hg'
Hg <- Hg.frame$Hg   # equivalent to 'Hg.frame[,1]' or `Hg.frame[,"Hg"]`
str( Hg) # now we have a vector
 is.na( Hg)
which( is.na( Hg))    # check if there are NAs, alt: 'any( is.na( Hg))' 

# str( read.csv('data/lemanHg.csv', header=FALSE))
            # Wrong way to import data. Result is a 'factor' not 'numeric'!
```

### R-Code (not in the script):
### Working with a vector 
```{r}
my.data <- c(2, 3, 5, 7,1)
my.data[1]
my.data[3]
my.data[my.data > 4]
```


### R-Code: 
### Example of creating ordinal and interval scales in R.
### (Lecture II)
```{r}
ordinal <- factor( c("male","female"))
ordinal[1] == ordinal[2]
#ordinal[1] > ordinal[2]  # warning ">" not meaningful for factors
interval <- c(2, 3)
interval[1] > interval[2] 
```


### R-Code: 
### A quantitative EDA of the lemanHg dataset. (Lecture III)
```{r}
mean( Hg)
mean( Hg, trim=0.1)
median( Hg)


c( mean=mean( Hg), tr.mean=mean( Hg, trim=0.1), median=median( Hg))
c( var=var( Hg), sd=sd( Hg), iqr=IQR( Hg))   # capital letters for IQR!
summary( Hg)              # min, max, quartiles and mean
range( Hg)                # min, max, but not the difference 
tail( sort( Hg))          # sorts and then list the 6 largest values 

## dat <- c(2, 15, 16, 32, 25, 10)   # see Figure 1.11
## emissionsource <- c('Air', 'Transp', 'Manufac', 'Electr', 'Deforest', 'Other')
## barplot( dat, names=emissionsource, ylab="Percent", las=2)
## barplot( cbind('2005'=dat), col=c(2,3,4,5,6,7), legend=emissionsource,
##       args.legend=list(bty='n'), ylab='Percent', xlim=c(0.2,4))
```



### R-Code/Figure: 
### Bar plots: (Lecture IV)
```{r}
dat <- c(2, 15, 16, 32, 25, 10)   # see Figure 1.11
emissionsource <- c('Air', 'Transp', 'Manufac', 'Electr', 'Deforest', 'Other')
barplot(dat)
barplot( dat, names=emissionsource, ylab="Percent")

################## (not in the lecture)
barplot( cbind('2005'=dat), col=c(2,3,4,5,6,7), legend=emissionsource,
      args.legend=list(bty='n'), ylab='Percent', xlim=c(0.2,4))
```      



### R-Code/Figure: 
### Histogram: (lecture V)
```{r}
par( mfrow=c(2, 2))
histout <- hist( Hg)    # default histogram 
hist( Hg, col=7, probability=TRUE, main="With 'smoothed density'")
lines( density( Hg))    # add smooth version of the histogram
abline( v=c(mean( Hg), median( Hg)), col=3:2, lty=2:3, lwd=2:3) 
hist( Hg, col=7, breaks=90, main="Too many bins")
hist( Hg, col=7, breaks=2, main="Too few bins")
str( histout[1:3])      # Contains essentially all information of histogram
```      






### R-Code/Figure: 
### Boxplots: (lecture VI)
```{r}
par( mfcol=c(1, 3)) # to have three plots next to each other. 
out <- boxplot( Hg, col="LightBlue", ylab="Hg")

out <- boxplot( Hg, col="LightBlue", ylab="Hg", outlty=1, outpch='') 

  # 'out' contains numeric values of the boxplot 
quantile( Hg, c(0.25, 0.75))  # compare with summary( Hg) and out["stats"]
IQR( Hg)                      # interquartile range  
quantile(Hg, 0.75) + 1.5 * IQR( Hg)           # upper boundary of the whisker
Hg[ quantile(Hg, 0.75) + 1.5 * IQR( Hg) < Hg] # points beyond the whisker

require(vioplot)                         # R package providing violin plots                
vioplot( Hg, col="Lightblue", ylab="Hg") 
```      





### R-Code/Figure: 
### QQplots: (lecture VII)
```{r}
qqnorm( Hg)            # QQplot with comparing with bell-shaped theoretical
qqline( Hg, col=2, main='')  # add read line 
theoQuant <- qchisq( ppoints( 293), df=5) # minor mystery for the moment
# hist( theoQuant, prob=TRUE); lines(density(theoQuant)) # convince yourself
qqplot( theoQuant, Hg, xlab="Theoretical quantiles")
# For 'chisq' some a priori knowledge was used, for 'df=5' minimal 
# trial and error was used.
qqline( Hg, distribution=function(p) qchisq( p, df=5), col=2)
``` 



### R-Code/Figure: 
### Barplots:
```{r}
par( mfcol=c(1,2))
dat2 <- c(2, 10, 12, 28, 26, 22) # source c2es.org
mat <- cbind( SWISS=dat, c2es.org=dat2)
barplot(mat, col=c(2,3,4,5,6,7), xlim=c(0.2,5), legend=emissionsource,
        args.legend=list(bty='n'), ylab='Percent', las=2)
barplot(mat, col=c(2,3,4,5,6,7), xlim=c(1,30), legend=emissionsource,
        args.legend=list(bty='n'), ylab='Percent', beside=TRUE, las=2)
```      


### R-Code (not in the script):
### Scatter plot for penguins data set
```{r}
require(palmerpenguins)
data("penguins") # the data is available from the package palmerpenguins
str(penguins)
pairs( penguins[,c(3:6)])
pairs( penguins[,c(3:6)], col=penguins$species)
```




### R-Code/Figure: 
### Histograms of the flipper length of the penguin dataset and further graphics
```{r}
require(palmerpenguins)
penguins <- read.csv(       # 'path.package()' provides local location of pkg    
	paste0( path.package('palmerpenguins'),'/extdata/penguins.csv'))
str(penguins, strict.width='cut')
summary(penguins[, c(3:6)])  # others variables can be summarized by 'table()'
table(penguins[,c(2,1)])     # tabulating species on each island

hist( penguins$flipper_length_mm, main='', xlab="Flipper length [mm]", col=7)
box()                        # rectangle around for similar view with others
with(penguins,  vioplot(flipper_length_mm[species=="Gentoo"], 
      flipper_length_mm[species=="Chinstrap"], flipper_length_mm[ 
      species=="Adelie"],  names=c("Gentoo", "Chinstrap", "Adelie"), 
      col=5:3, xlab="Flipper length [mm]", horizontal=TRUE, las=1))
mosaicplot( table( penguins[,c(2,1)]), col=3:5, main='', cex.axis=.75, las=1)
upper.panel <- function( x,y, ...)   # see '?pairs' for a better and more 
  points( x, y, col=as.numeric(penguins$species)+2,...) # complete way
lower.panel <- function( x,y, ...) 
  points( x, y, col=as.numeric(penguins$sex))

```



### R-Code/Figure:
### Parallel coordinate plot of the swiss dataset:
```{r}
dim( swiss)   # in   package:datasets, available without the need to load.
# str( swiss, strict.width='cut')    # or even:
# head( swiss);  summary( swiss)
require( MASS)    # package providing the function `parcoord()`
parcoord( swiss, col=2-(swiss$Catholic<40) + (swiss$Catholic>60))
```


########################################################################
# Reinhard Furrer, Spring 2023