---
title: "02 Random Variable"
author: "Reinhard Furrer with contributions of various others"
date: "February 26, 2024"
output:
  pdf_document: 
    latex_engine: xelatex
    highlight: tango
  word_document: default
subtitle: 'STA120: Introduction to Statistics Spring semester 2023 IMath, University
  of Zurich, Switzerland'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
 
```

### Check the following functions to compute the density/pmf (d), the cumulative distribution function (p), the quantile function (q), or to sample from a distribution (r).

?pnorm, qnorm, dnorm, rnorm ?pbinon, qbinom, dbinom, rbinom ?ppois, qpois, dpois, rpois  θ

###Examples of densities, quantiles and cdf \### Lecture I

```{r}
pnorm(0)
pnorm(0, mean = 4, sd = 4) #in R, one uses sd and not the variance as the parameter
pnorm(0, 4, 4)

qnorm(0.5)
qnorm(0.999)

pbinom(q = 0, size = 2, prob = 0.5)

dbinom(x = 1, size = 2, prob = 0.8)
```

### Ploting: pmf, cdf, density. Lecture II

```{r}
# a pmf:
x <- c(0:10)
p <- dbinom(x = x, prob = 0.5, size = 10)
plot(x, p)
plot(x, p, type = "h", lwd = 5, col = "Darkred")

# a cdf:
#first try:
cdf <- pbinom(q = x, prob = 0.5, size = 10)
plot(x, cdf) # only in the defined points, we would like to see the whole cdf

#second try:
x2 <- seq(from = -1, to = 11, by = 0.01)
cdf2 <- pbinom(q = x2, prob = 0.5, size = 10)
plot(x2, cdf2, lwd = 1, t = "l", col = "Darkmagenta") # that does not look like a function

#third try:
plot( stepfun( x, c(0,cdf))) # that still does not look like a function

plot( stepfun( x, c(0,cdf)), ylab=bquote(F[X](x)), verticals=FALSE,  main='', pch=20)
# And finally a proper plot of a cdf

# a density:
x3 <- seq(from = -5, to = 5, by = 0.01)
d <- dnorm(x = x3, mean = 1, sd = 0.75)
plot(x3, d, type = "l", col = "mistyrose3", lwd = 3)
```

### R-Code/Figure:

### Pmf and distribution function of a binomial distribution

### Lecture III

```{r}
par( mfrow=c(2, 2))

plot( 0:12, dbinom(0:12, size=12, prob=1/3)) # the easiest version

plot( 0:12, dbinom(0:12, size=12, prob=1/3), type='h',
      xlab='x', ylab=bquote(f[X](x)))  

# or a bit more elegant and a bit more complex
plot( stepfun( 0:12, pbinom(-1:12, size=12, prob=1/3)), 
           ylab=bquote(F[X](x)), verticals=FALSE,  main='', pch=20)

set.seed( 14)     # same results if the following lines are evaluated again
print( x10 <- rbinom(10, size=12, prob=1/3 ))    # printing the 10 draws
barplot( table( factor(x10, levels=0:12)))       # barplot of the 10 draws
x100 <- rbinom(100, size=12, prob=1/3 )          # 100 draws
barplot( table( factor(x100, levels=0:12))/100)  # frequency barplot 
sum(x100==5) 

hist(x100, freq = FALSE, breaks = c(0:12))
lines(density(x100))

qqnorm(x100)
# how many times five face cards?
## alternative way for the cdf:
# plot( c(-.5,0,1,1.5), c(0,0,1,1), type='l', xlab='x', ylab=bquote(F[X](x)))
```

### R-Code/Figure:

### Number of earthquakes and pmf of a Poisson distribution

### Lecture IV

```{r}
mag4 <- c(2,0,0,2,5,0,1,4,1,2,1,3,3,2,2,3,0,0,3,1,2,2,2,4,3) # data
hist( mag4, breaks=seq(-0.5, to=10), prob=TRUE, main="", col='gray')
points( 0:10, dpois(0:10, lambda=2), pch=19, col=4)  # pmf of X~Pois(2)
```

############################################################################## 

#### What would be a pseudocode for pmf and cdf of the sum of the roll of two dice?

############################################################################## 

### Function outer()

```{r}
outer(1:6,1:6,"+")
outer(1:6,1:6)
outer(1:6,1:6,"/")
```

### Function plot.ecdf() - Plot empirical cumulative distribution function

```{r}
plot.ecdf(c(1,1,1,2,2,3))
```

### R-Code/Figure:

### pmf and cdf of the sum of the roll of two dice:

```{r}
par( mfrow=c(1, 2))
plot.ecdf( outer(1:6, 1:6, "+"),     # generating all possible outcomes
           ylab=bquote(F[X](x)), main='', pch=20)  # `bquote` for subscripts
x <- 2:12               #  possible outcomes
p <- c(1:6, 5:1)/36     #  corresponding probabilities
plot( x, p, type='h', xlim=c(1,13), ylim=c(0, .2),
     xlab=bquote(x[i]), ylab=bquote(p[i]==f[X](x[i])))
points( x, p, pch = 20) # adding points for clarity
```

### R-Code/Figure:

### Density and distribution function of the uniform continuous distribution

```{r}
par( mfrow=c(1, 2))
plot( c(-0.5, 0, NA, 0, 1, NA, 1, 1.5), c(0, 0, NA, 1, 1, NA, 0, 0),  
      type='l', xlab='x', ylab=bquote(f[X](x)))  
# curve(dunif( x), -0.5, 1.5) # does not emphasize the discontinuity!!
curve(punif( x), -0.5, 1.5)
## alternative way for the cdf:
# plot( c(-.5,0,1,1.5), c(0,0,1,1), type='l', xlab='x', ylab=bquote(F[X](x)))
```

############################################################################ 

# Construction of Figure

# This code is not shown in the script as an R-Chunk and is for illustration

# purposes only.

```{r}
adens <- function( x, y, ...) {
	polygon(c(x[1], x, x[length(x)]), c(0,y,0),..., border=NA)
}

par( mfcol=c(1,2), mai=c(.3,.1,.1,.1), cex=.75)
h <- seq( -3.5, 3.5, l=100)
col <- c( rgb(1,0,0,1), rgb(0,1,0,1), rgb(0,0,1,1))
plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3.2, 3.2))
adens( h[h< 2],  dnorm( h[h< 2]), col=col[1], density=30) 
adens( h[h< 1],  dnorm( h[h< 1]), col=col[2], density=35, angle=90) 
adens( h[h< 0],  dnorm( h[h< 0]), col=col[3], density=20, angle=-45) 
text(3.1, .37, "P(Z<0)=50.0%", col=col[3], adj=1)
text(3.1, .32, "P(Z<1)=84.1%", col=col[2], adj=1)
text(3.1, .27, "P(Z<2)=97.7%", col=col[1], adj=1)
round( pnorm(0:2),3)

plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3.2, 3.2))
adens( h[h< 3 & h > -3],  dnorm( h[h< 3 & h> -3]), col=col[1], density=30) 
adens( h[h< 2 & h > -2],  dnorm( h[h< 2 & h> -2]), col=col[2], density=35, angle=90) 
adens( h[h< 1 & h > -1],  dnorm( h[h< 1 & h> -1]), col=col[3], density=20, angle=-45) 
text(3.2, .37, "P(-1<Z<1)=68.3%", col=col[3], adj=1)
text(3.2, .32, "P(-2<Z<2)=95.4%", col=col[2], adj=1)
text(3.2, .27, "P(-3<Z<3)=99.7%", col=col[1], adj=1)
round( pnorm(1:3)-pnorm(-1:-3),3)
```

############################################################################ 

### R-Code/Figure:

### pdf, cdf and quantile function of the Gaussian distribution

```{r}
plot( dnorm, -3, 3, ylim=c(-.5,2), xlim=c(-2.6,2.6)) 
abline( c(0, 1), h=c(0,1), v=c(0,1), col='gray') # diag and horizontal lines
plot( pnorm, -3, 3, col=3, add=TRUE) 
plot( qnorm, 0, 1, col=4, add=TRUE)
```

######################################################################## 

# Reinhard Furrer, Spring 2023
