---
title: "03 Functions of random variables"
subtitle: "STA120: Introduction to Statistics
         Spring semester 2024
         IMath, University of Zurich, Switzerland"
author: "Reinhard Furrer with contributions of various others"
date: "February 29, 2024"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###Example 1 set.seed(). lecture I
```{r}
x <- rnorm(n = 1, mean = 0, sd = 1)
# equivalent to rnorm(1,0,1), or rnorm(1)
print(x)
```


### Example 2 set.seed(). Lecture II
```{r}
set.seed(1)
x <- rnorm(n = 1, mean = 0, sd = 1)
# equivalent to rnorm(1,0,1), or rnorm(1)
print(x)

y <- rnorm(n = 1, mean = 0, sd = 1)
# equivalent to rnorm(1,0,1), or rnorm(1)
print(y)
```


### Distribution of a sum of independent normal RVs. Lecture III
```{r}
set.seed(1)
n <- 10000
x <- rnorm(n, 2, 4) # I choose arbitrary values for mu and sigma
y <- rnorm(n, 1, 1)
z <- rnorm(n, -3, 2)
obs <- x+y+z

grid <- seq(-15, 15, length.out = 1000)

hist(obs, probability = TRUE)
lines(density(obs), col = "chartreuse4")
lines(grid, dnorm(grid, mean = 2+1-3,sd = sqrt(16+1+4)), col = "darkmagenta") 
#Var(X+Y+Z)=(because X,Y,Z independent)
#=Var(X)+Var(Y)+Var(Z) = 4^2 +1^2 +2^2
```

### Chi square distribution. Lecture IV
```{r}
set.seed(1)
n <- 100000
x <- rnorm(n, 0, 1)
grid <- seq(0, 5, length.out = 10000)
x2 <- x*x 
hist(x2, probability = TRUE, ylim = c(0,1), xlim = c(0.1,10))
lines(density(x2), col = "chartreuse4")
lines(grid, dchisq(grid, df = 1), col = "darkmagenta")
```

### Chi square distribution with 5 degrees of freedom. Lecture V.
```{r}
set.seed(1)
n <- 100000
x2 <- numeric(n)
grid <- seq(0, 10, length.out = 10000)
for (i in 1:n){
  xs <- rnorm(5,0,1)
  x2[i] <- sum(xs*xs) 
}
hist(x2, probability = TRUE, ylim = c(0,1), xlim = c(0.1,10))
lines(density(x2), col = "chartreuse4")
lines(grid, dchisq(grid, df = 5), col = "darkmagenta")
```

### Distribution of standardized means of normal RVs.
### Variance not known. t-distribution
### Lecture VI
```{r}
set.seed(1)
n <- 10000

means <- numeric(n)  # prepare space for storing the means
st.x <- numeric(n)
for (i in c(1:n)){
  x <-rnorm(3, 2, 1) #sample from X1, X2, X3 iid N(2,1)
  means[i] <- mean(x)
  st.x[i] <- (means[i]-2)/sqrt(var(x)/3)
}

grid <- seq(-3,3, length.out = 1000)

hist(st.x, probability = TRUE,breaks = 200, ylim=c(0,1), xlim=c(-3,3))
lines(density(st.x), col = "chartreuse4")
lines(grid, dnorm(grid, mean = 0,sd = 1), col = "darkmagenta") 
lines(grid, dt(grid, df=2), col = "red") 
```



### R-Code/Figure: 
### Densities of the Chi-square distribution:
```{r}
x <- seq( 0, to=50, length=150)               # x-values for which we plot  
plot(x, dchisq( x, df=1), type='l', ylab='Density')  # plot first density
for (i in 1:6)                                # different degrees of freedom 
    lines( x, dchisq(x, df=2^i), col=i+1)     
legend( "topright", legend=2^(0:6), col=1:7, lty=1, bty="n")
```     
      



### R-Code/Figure: 
### Densities of the t-distribution:
```{r}
x <- seq( -3, to=3, length=100)              # x-values for which we plot  
plot( x, dnorm(x), type='l', ylab='Density') # Gaussian density as reference
for (i in 0:6) 
    lines( x, dt(x, df=2^i), col=i+2)        # t-densities with different dofs
legend( "topright", legend=2^(0:6), col=2:8, lty=1, bty="n")
```      
      



### R-Code: 
### sample variances of the t-distribution with one degree of freedom:
```{r}
set.seed( 14) 
tmp <- rt( 1000, df=1)   # 1000 realizations
var( tmp)                # variance is huge!!
sort( tmp)[1:7]      # many "large" values, but 2 exceptionally large
sort( tmp, decreasing=TRUE)[1:7] 
```

### F-distribution. Lecture VII
```{r}
M <- 100000
m <- 15
n <- 13
Ws <- numeric(100000)
for (i in 1:M){
  x <- rnorm(m)
  y <- rnorm(n)
  Ws[i] <- (sum(x*x)/m)/(sum(y*y)/n)
}
grid <- seq(-0.5, 5, length.out = 10000)
hist(Ws, probability = TRUE, breaks = 200, xlim=c(0,5))

lines(density(Ws), col = "chartreuse4")
lines(grid, df(grid, df1=m,df2=n), col = "darkmagenta") 
 
```

### Loops. Lecture VIII
```{r}
means <- numeric(100)
for(i in 1:100){
  nsample <- rnorm(n = 100, mean = 0, sd = 1)
  means[i] <- mean(nsample)
}

hist(means)
```



### R-Code/Figure: 
### Densities of the F-distribution:
```{r}
x <- seq(0, to=4, length=500)            # x-values for which we plot  
df1 <- c( 1, 2, 5, 10, 50, 50, 250)      # dof for the numerator
df2 <- c( 1, 50, 10, 50, 50, 250, 250)   # dof for the denumerator
plot( x, df( x, df1=1, df2=1), type='l', ylab='Density') 
for (i in 2:length(df1))
    lines( x, df(x, df1=df1[i], df2=df2[i]), col=i) 
legend( "topright", col=1:7, lty=1, bty="n", 
        legend=parse(text=paste0("F['",df1,",",df2,"']")))
```      

########################################################################
# Reinhard Furrer, Spring 2023