---
title: "Lecture05, FS24 STA120: Introduction to Statistics"
author: "Reinhard Furrer with contributions of various others"
date: "March 16, 2024"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### An example, that the true mean may not be within the CI for this parameter
```{r}
set.seed(14)
x <- rnorm(10,0,1)
CI <- mean(x)+c(qnorm(c(0.025, 0.975), sd =1 ))/sqrt(10)
print(CI)
```


### An example showing that the likelihood may be greater than 1
```{r}
set.seed(1)
grid <- seq(-2,2, length.out = 1000)

x <- rnorm(1,0,0.1)
plot(grid, dnorm(grid, 0, 0.1), type = "l", col="darkorchid3")
abline(v = x)
abline(h =dnorm(x, 0, 0.1))
print(x)
print(dnorm(x, 0, 0.1))

```


### t-test for one sample. Here - randomly selected 10 numbers following a normal distribution, rounded to integers. t-test to compare their mean to 100.

```{r}
options(digits = 5)
set.seed(10)
x <- round(rnorm(10,110,8 ))
x
t.test(x, mu =100)
```


### An illustration to the test above.
```{r}
tobs <- (mean(x)-100)/(sd(x)/(sqrt(10)))
tobs
grid <- seq(-10,10,length.out = 1000)
plot(grid, dt(grid, df = 9), type = "l", col ="darkorchid3")
abline(v = tobs)
abline(v = -tobs)
abline(v = qt(0.025, df =9), col ="red")
abline(v = qt(0.975, df =9), col ="red")
pval <- 2*(1- pt(tobs, df = 9))
pval

```

############################################################################ 
############################################################################ 
#  This code is not shown in the script as an R-Chunk and is for illustration
#  purposes only.
# 
# Figure: 
# Illustration of the p-values
```{r}
adens <- function( x, y, ...) {
  polygon(c(x[1], x, x[length(x)]), c(0,y,0),...,border=NA)
}

obs <- 1.8
h <- seq( -3.3, 3.5, l=250)
plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3,3.2))
text( -2, .3, bquote(H[0]), cex=1.3)
rug(obs, col = "DarkGreen", ticksize = .4, lwd=3)
text( obs+.6, .2,"Observation", cex=1.1, col="DarkGreen")
adens( h[h< -obs],  dnorm( h[h< -obs]), col="DarkGreen") 
adens( h[h>  obs],  dnorm( h[h> obs]), col="DarkGreen") 
text( 3.2, .34, paste0("p-value=",2*round(1-pnorm(obs),3),"\n (two-sided)"), adj=1, cex=1.1)

plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3,3.2))
text( -2, .3, bquote(H[0]), cex=1.3)
rug(obs, col = "DarkGreen", ticksize = .4, lwd=3)
text( obs+.6, .2,"Observation", cex=1.1, col="DarkGreen")
adens( h[h>  obs],  dnorm( h[h> obs]), col="DarkGreen") 
text( 3.2, .34, paste0("p-value=",round(1-pnorm(obs),3),"\n (one-sided)"), adj=1, cex=1.1)
```


### Figure:
### Critical values and rejection region.
```{r}
quant <-  qnorm(.975) #  is approx 1.96
h <- seq( -3.3, 3.5, l=250)
plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3,3.2))
text( -2, .3, bquote(H[0]), cex=1.3)
text( c(-2.4, 2.4), .2,"Critical value", cex=1.1, col=2)
rug( c(-quant, quant), col="Brown", ticksize = .4, lwd=3)
lines( c(quant, 4), -c(.01,.01), col="Orange", lwd=5)
lines( -c(4, quant), -c(.01,.01), col="Orange", lwd=5)

quant <-  qnorm(.95)
plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3,3.2))
text( -2, .3, bquote(H[0]), cex=1.3)
text( quant+.7, .2,"Critical value", cex=1.1, col=2)
rug( quant, col="Brown", ticksize = .4, lwd=3)
lines( c(quant, 4), -c(.01,.01), col="Orange", lwd=5)
```


# Construction of Figure 4.4: Type I and Type II errors.
#  This code is not shown in the scrip as an R-Chunk.
```{r}
h <- seq( -4, 8, l=700)
quant <- qnorm(.95)  #  which is approx 1.64

plot( h, dnorm(h), type='l', xlab='',ylab='',yaxt='n', xlim=c(-3,7))
adens( h[h>quant],  dnorm( h[h>quant]), col=2) 
adens( h[h< -quant],  dnorm( h[h< -quant]), col=2) 
lines( h, dnorm(h,mean=2))
adens( h[h<= quant & h>-quant],  dnorm(h[h<= quant & h>-quant], mean=2), col=4) 
text( -2, .3, bquote(H[0]), cex=1.3)
text(  4, .3, bquote(H[1]), cex=1.3)

plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3,7))
adens( h[h>quant],  dnorm( h[h>quant]), col=2) 
adens( h[h< -quant],  dnorm( h[h< -quant]), col=2) 
lines( h, dnorm(h,mean=4))
adens( h[h<= quant],  dnorm( h[h<= quant],mean=4), col=4) 
text( -2, .3, bquote(H[0]),cex=1.3)
text(  6, .3, bquote(H[1]),cex=1.3)

plot( h, dnorm(h), type='l', xlab='',ylab='',yaxt='n', xlim=c(-3,7))
adens( h[h>quant],  dnorm( h[h>quant]), col=2) 
lines( h, dnorm(h,mean=2))
adens( h[h<= quant],  dnorm(h[h<= quant],mean=2), col=4) 
text( -2, .3, bquote(H[0]), cex=1.3)
text(  4, .3, bquote(H[1]), cex=1.3)

plot( h, dnorm(h), type='l', xlab='', ylab='', yaxt='n', xlim=c(-3,7))
adens( h[h>quant],  dnorm( h[h>quant]), col=2) 
lines( h, dnorm(h,mean=4))
adens( h[h<= quant],  dnorm( h[h<= quant],mean=4), col=4) 
text( -2, .3, bquote(H[0]),cex=1.3)
text(  6, .3, bquote(H[1]),cex=1.3)
```

# End Figures.
############################################################################ 
############################################################################ 


### R-Code/Figure
### Power-Kurve:
```{r}
alpha <- 0.05                       # significance level
mu0 <- 0                            # mean under H_0
mu1 <- seq(-1.5, to=5, by=0.1)      # mean under H_1
power_onesided <- 1-pnorm( qnorm(1-alpha, mean=mu0), mean=mu1)
power_twosided <- pnorm( qnorm(alpha/2, mean=mu0), mean=mu1) +
    pnorm( qnorm(1-alpha/2, mean=mu0), mean=mu1, lower.tail=FALSE)
plot( mu1, power_onesided, type='l', ylim=c(0,1), xlim=c(-1, 4.25), las=1,
     xlab=bquote(mu[1]-mu[0]),  ylab="Power", col=4, yaxs='i', lwd=1) 
axis(2, at=alpha, labels='')        # adding level
axis(2, at=1.4*alpha, labels=bquote(alpha), las=1, adj=0, tick=FALSE)
lines( mu1, power_twosided, lty=2)  # power curve for two-sided test
abline( h=alpha, col='gray')        # significance level
abline( v=c(2, 4), lwd=2, col=3)    # values from figure 4.3
#       
```  


```{r}
podo <- read.csv("http://user.math.uzh.ch/furrer/download/intro2stat/podo.csv")

apply( podo, 2, function(x) length(unique(x)) ) # 4 visits, 2 barns

PDHmean <- podo$PDHmean[podo$Visit==13]
length( PDHmean)
print( me <- mean( PDHmean))
print( se <- sd( PDHmean))

```

### Example 4.5:
```{r}
PDHmean
mean(PDHmean)
sd(PDHmean)
n <- length(PDHmean)
tobs <- abs( mean(PDHmean) - 3.333)/sd(PDHmean)*sqrt(n)
qt(0.975, n)
2*(1-pt( tobs, n-1))  # equivalent to 2*pt( -tobs, n-1)

print( out <- t.test( PDHmean, mu=3.333))   # print the result of the test
out$p.val                                   # printing only the p-value
```

### Choosing data for the paired test
```{r}
podoV1V13 <- podo[podo$Visit %in% c(1,13),] # select visits from 2016 and 2017
PDHmean2 <- matrix(podoV1V13$PDHmean[order(podoV1V13$ID)], ncol=2, byrow=TRUE)
t.test( PDHmean2[,2], PDHmean2[,1], paired=TRUE)
```


### Example 4.6:
```{r}
PDHmeanB1 <- podoV1V13$PDHmean[podoV1V13$Barn==1]     
PDHmeanB2 <- podoV1V13$PDHmean[podoV1V13$Barn==2]
mean(PDHmeanB1)
mean(PDHmeanB2)
nx <- length(PDHmeanB1)
ny <- length(PDHmeanB2)
s_p <- sqrt( 1/(nx+ny-2)*( (nx-1)* var(PDHmeanB1) + 
                           (ny-1)* var(PDHmeanB2))) 
tobs <- abs( mean(PDHmeanB1) - mean(PDHmeanB2) )/ 
              s_p*sqrt(nx*ny/(nx+ny))
qt(0.975, nx+ny-2)
2*(1-pt( tobs, nx+ny-2))  # equivalent to 2*pt( -tobs, nx+ny-2)

t.test( PDHmeanB1, PDHmeanB2, var.equal=TRUE)
```


### Example 4.7:
```{r}
podoV1V13 <- podo[podo$Visit %in% c(1,13),]
PDHmean2 <- matrix(podoV1V13$PDHmean[order(podoV1V13$ID)], ncol=2, byrow=TRUE)
difference <- PDHmean2[,2] - PDHmean2[,1]
mean( difference)
sd( difference)
n <- length(difference)
tobs <- abs( mean(difference))/sd(difference)*sqrt(n)
qt(0.975, n-1)
2*(1-pt( tobs, n-1)) 

podoV1V13 <- podo[podo$Visit %in% c(1,13),] # select visits from 2016 and 2017
PDHmean2 <- matrix(podoV1V13$PDHmean[order(podoV1V13$ID)], ncol=2, byrow=TRUE)
t.test( PDHmean2[,2], PDHmean2[,1], paired=TRUE)
# Same result as with 
# t.test( PDHmean2[,2] - PDHmean2[,1])



### R-Code
### Comparing two variances:
var.test( PDHmeanB1, PDHmeanB2)
```
########################################################################
# Reinhard Furrer, Spring 2023